---
title: "Alpha diversity"
output: 
  pdf_document: 
    keep_tex: yes
---


```{r, setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, eval=FALSE, tidy.opts=list(width.cutoff=20), tidy=TRUE, wrap=TRUE)

```

Plots and statistics for alpha-diversity.

Joint data frame for metadata and alpha diversity.
```{r}
VFLA_alpha_data <-metadata %>% 
  left_join(VFLA_shannon, by=c("SampleID"="SampleID")) %>%
  left_join(VFLA_faith_pd, by=c("SampleID"="SampleID")) %>% 
  mutate(Ring = factor(
      Ring, 
      levels = c("Ring_1", "Ring_3", "Ring_8", "Ring_4", "Ring_7", "Ring_12"),
      labels = c("Pen 1", "Pen 3", "Pen 8", "Pen 4", "Pen 7", "Pen 12")
    ),
        Type = factor(Type, levels = c("Plastic", "Seaweed", "Skin", "Gills")))%>% 
  filter(Shelter != "Growth")
```


Alpha diversity split by cages and sample type with facet on shelter type
```{r}

faith_pd_plot<-ggplot(VFLA_alpha_data, aes(x = Ring, y = faith_pd, fill = Type))+
  geom_boxplot()+
  xlab("Pen number")+
  ylab("Faith's PD")+
  scale_fill_brewer(palette = "Paired") +
  facet_wrap(~Shelter1, scales = "free_x", dir = "h", ncol = 2) +
  theme_classic() +
  theme(text = element_text(size = 13))

shannon_plot<-ggplot(VFLA_alpha_data, aes(x = Ring, y = shannon_entropy, fill = Type))+
  geom_boxplot()+
  xlab("Pen number")+
  ylab("Shannon")+
  scale_fill_brewer(palette = "Paired")+
  facet_wrap(~Shelter1, scales = "free_x") +
  theme_classic() +
  theme(text = element_text(size = 13))
  

```

Alpha diversity split by sample type and sampling time point with facet on shelter type
```{r}
shannon_all <- ggplot(VFLA_alpha_data, aes(x = Type1, y = shannon_entropy, fill = Cycle))+ 
  geom_boxplot()+
  facet_grid(Shelter1~.)+
  scale_fill_brewer(palette = "OrRd")+
  labs(x = "Type", y = "Shannon", fill = "Sampling point")+
  theme_classic()+
  theme(text = element_text(size = 13))

pd_all <- ggplot(VFLA_alpha_data, aes(x = Type1, y = faith_pd, fill = Cycle))+ 
  geom_boxplot()+
  facet_grid(Shelter1~.)+
  scale_fill_brewer(palette = "OrRd")+
  labs(x = "Type", y = "Faith's PD", fill = "Sampling point")+
  theme_classic()+
  theme(text = element_text(size = 13))

```

```{r, fig.height=10, fig.width=12}
ggarrange(shannon_plot, faith_pd_plot, shannon_all,pd_all,nrow = 2, ncol = 2, labels = c("a)","b)","c)","d)"))
ggsave("alpha_diversity_plots.tiff",dpi=300)

```


Comparisons of sample types. Relevel is used to get a different base level, so that all comparisons can be made
```{r}
VFLA_alpha_data$Type <- relevel(factor(VFLA_alpha_data$Type),ref="Gills")
VFLA_type_faithmod <-lmer(faith_pd~Type+(1|Ring),data=VFLA_alpha_data)
summary(VFLA_type_faithmod)

VFLA_type_shannonmod <-lmer(shannon_entropy~Type+(1|Ring),data=VFLA_alpha_data)
summary(VFLA_type_shannonmod)

```


Gills and Skin in plastic and seaweed shelters
```{r}
VFLA_alpha_data$Shelter1 <- relevel(factor(VFLA_alpha_data$Shelter1),ref="Seaweed")

VFLA_faithmod_shelter_gills <-lmer(faith_pd~Shelter1+(1|Ring),data=subset(VFLA_alpha_data, Type == "Gills"))
summary(VFLA_faithmod_shelter_gills)

VFLA_shannonmod_shelter_gills <-lmer(shannon_entropy~Shelter1+(1|Ring),data=subset(VFLA_alpha_data, Type == "Gills"))
summary(VFLA_shannonmod_shelter_gills)

VFLA_faithmod_shelter_skin <-lmer(faith_pd~Shelter1+(1|Ring),data=subset(VFLA_alpha_data, Type == "Skin"))
summary(VFLA_faithmod_shelter_skin)

VFLA_shannonmod_shelter_skin <-lmer(shannon_entropy~Shelter1+(1|Ring),data=subset(VFLA_alpha_data, Type == "Skin"))
summary(VFLA_shannonmod_shelter_skin)

```

Change over time
```{r}
VFLA_alpha_data$Cycle <- relevel(factor(VFLA_alpha_data$Cycle),ref="A")

VFLA_faithmod_time <-lmer(faith_pd~Cycle+(1|Type),data=VFLA_alpha_data)
summary(VFLA_faithmod_time)

VFLA_shannonmod_time <-lmer(shannon_entropy~Cycle+(1|Type),data=VFLA_alpha_data)
summary(VFLA_shannonmod_time)

```


